---
title: "TU R Group Week 1 Regression"
author: "Ann Marie Finley"
date: "`r format(Sys.Date())`"
output:
   html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    css: style.css
---

```{r setup, include=FALSE }
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/', echo=TRUE, tidy=TRUE, message=F, warning=F, cache=T)
```


```{r, echo=F}
library(tidyverse)
library(reshape2)
library(dendextend)
library(corrplot)
library(tibble)
jamie.theme <- theme_bw() + theme(axis.line = element_line(colour = "black"), panel.grid.minor = element_blank(), panel.grid.major = element_blank(), panel.border = element_blank(), panel.background = element_blank(), legend.title= element_blank())
```

# Read in data
Look at structure, column names, headers and first ten rows
```{r}
wone <- read.csv("cursing2.csv" , header = T)
str(wone)
names(wone)
head(wone, n=10)
```

# Basic Scatterplot 
Length x Tabooness
```{r}
plot(wone$Nlett ~ wone$Taboo, col = "navy")
```

# Fancy Scatterplot
Loess Function
```{r, fig.height=6, fig.width=8}
ggplot(wone, aes(Taboo, Nlett)) + geom_point(shape = 2, size = 1, color = "navy", 
    alpha = 0.7) + stat_smooth(method = loess, color = "pink") + xlab("Taboo") + 
    ylab("Nlett") + scale_x_continuous(breaks = seq(0, 7, by = 1)) + 
    scale_y_continuous(breaks = seq(0, 7, by = 1)) + jamie.theme
```

# Regression 
Do number of letters and valence predict word tabooness?
```{r}
reg1 <- lm(wone$Taboo ~ wone$Nlett + wone$Valence)
summary(reg1)
```

Tabooness = 3.33 + Nlett(-0.02) + Valence(-0.31)

Number of letters was not a significant predictor of word tabooness. For every one unit increase in valence, there is a 0.31 decrease in word tabooness.
